proof-of-concept_mods

### 2023-11-03 proof-of-concept_mods.Rmd
plots, OHBM abstract analyses

Convert sigma to variance by correcting for log-link function (sex's effect on variance = exp(sigma) - exp(log(sigma) - beta sex)) -> exp(beta_sex).

```{r}
sigma.df <- sum.df %>%
  dplyr::filter(parameter == "sigma") %>%
  mutate(var_est = exp(estimate),
         )

@@

@@

@@

max.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_diff = max_centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=max_centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```
Other methods are fairly consistently underestimating male centiles and overestimating female centiles.

@@

mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_diff = centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```
T test of mean diff. between sex w/in each dataset
```{r}
t.test.by.sex <- mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

@@

@@

>>>>

Convert sigma to variance by correcting for log-link function (sex's effect on variance = exp(sigma) - exp(log(sigma) - beta sex)) -> exp(beta_sex).
Also getting mean variance.

```{r}
sigma.df <- sum.df %>%
  dplyr::filter(parameter == "sigma") %>%
  mutate(var_est = exp(estimate),
         var_mean = exp(moment.mean),
         beta_ratio = var_est/var_mean)

@@

```{r, fig.height=10, fig.width=8}
proportion_beta_plot <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = beta_ratio)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=15) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Proportion of effect of male sex on variance")
```

@@

##### Sex-Biases

@@

max.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_diff = max_centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=max_centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```

Other methods are fairly consistently underestimating male centiles and overestimating female centiles.

@@

mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_diff = centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```

T test of mean diff. between sex w/in each dataset
```{r}
t.test.by.sex <- mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

@@

Test for differences in distribution
```{r}
summary(aov(z_abs.diff ~ dataset, data=mean.z.long))

attach(mean.z.long)
pairwise.t.test(x=z_abs.diff, g=dataset, p.adj = "fdr")
detach()
```

##### Sex-Biases

Test for sex-specific differences in mean z-score offset

```{r}
z_diffs2 <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ . - first(.), .names = "diff_{.col}")) %>%
  ungroup()

mean.z.diffs.by.sex2 <- z_diffs2 %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="z_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno = gsub(".z", "", pheno)) %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))
```
t-test
```{r}
t.test.by.sex_z <- mean.z.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(z_diff ~ sex)$statistic,
            p_value = t.test(z_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

t.test.by.sex_z
```

Same results as centiles.

@@

#### OHBM Abstract

rerunning stats above w/o ref-site combat versions (to account for different FDR correction) per Andrew's suggestion

Test for differences in centile offset distributions
```{r}
mean.diffs.long_filt <- mean.diffs.long %>%
  filter(!str_detect(dataset, "_refA"))

summary(aov(centile_abs.diff ~ dataset, data=mean.diffs.long_filt))

attach(mean.diffs.long_filt)
pairwise.t.test(x=centile_abs.diff, g=dataset, p.adj = "fdr")
detach()
```

Testing for differences in Z offset distributions

```{r}
mean.z.long_filt <- mean.z.long %>%
  filter(!str_detect(dataset, "_refA"))

summary(aov(z_abs.diff ~ dataset, data=mean.z.long_filt))

attach(mean.z.long_filt)
pairwise.t.test(x=z_abs.diff, g=dataset, p.adj = "fdr")
detach()
```

Test for sex-biases in centile offsets

Test for sex-biases

```{r}
mean.z.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  filter(!str_detect(dataset, "_refA")) %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(z_diff ~ sex)$statistic,
            p_value = t.test(z_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  filter(!str_detect(dataset, "_refA")) %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))
```

##### Plots

Beta weight? or proportion?

```{r}
beta.plot2 <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss"))) %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = var_est)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Beta weight for effect of male sex on variance")

beta.plot2
```

```{r}
proportion_beta_plot2 <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss"))) %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = beta_ratio)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=15) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Proportion of effect of male sex on variance")

proportion_beta_plot2
```

Centile plot

```{r}
cent.diffs.plot2 <- mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  filter(!str_detect(dataset, "_refA")) %>%
  mutate(centile_abs.diff = centile_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_histogram(aes(x=centile_abs.diff, fill=dataset), color="black", alpha=0.4, position="identity") +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~pheno_cat) +
  #paletteer::scale_fill_paletteer_d("Polychrome::light") +
  ggtitle("Mean absolute differences in centile scores across \nComBat processing pipelines for each phenotype")

cent.diffs.plot2
```
